---
title: "Tutorial 1a: Exploring Pharmacological Data with the `rawPharmacoData` Dataset"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r echo=FALSE}
knitr::opts_chunk$set(cache=FALSE)
```

## Introduction

Probably the most important step of analyzing datasets is to actually understand 
the data. This process is crucial to know what kind of questions we can 
answer with it. This tutorial has code that will help guiding 
you through this process with the `rawPharmacoData` dataset.
Make sure you understand the experimental design of the
two studies well and try to link each variable to this experimental design.
Also, make sure you understand what each *R* command is doing.
Feel free to hack the code!

When it makes sense, we include examples for answering the question using
both base R and the tidyverse packages. There's usually more than one way of
doing things in R!

If you have any question about the code, ask one of the mentors.
Also remember that [Google search](https://www.google.com) is one of the most important tools
for data science. 

## Setup Workspace

We start by loading the tidyverse family of packages.

```{r}
library(tidyverse)
```

There are [several pre-defined themes for plotting with `ggplot2`](https://ggplot2.tidyverse.org/reference/ggtheme.html). 
While the default "`theme_gray`" is nice, we will set the default to "`theme_bw`" using the `theme_set` function.

```{r}
theme_set(theme_bw())
```

## Load Raw Dataset

Let's start by loading the `RDS` file containing the raw pharmacological data.

```{r read-rds}
pharmacoData <- readRDS(file.path("..", "data", "rawPharmacoData.rds"))
```

## Exploratory Analysis

We can take a quick peek at the data using the `head` and `str` functions.
What kind of variables are in the data? Are these variables numerical and/or categorical?
What does each column represent?

```{r peek-data}
head(pharmacoData)
str(pharmacoData)
```

Next, we can count the number of drugs and cell lines in the dataset.

```{r count-uniq}
## using base R
length(unique(pharmacoData$cellLine))
length(unique(pharmacoData$drug))

## with the tidyverse
pharmacoData %>%
    summarize(nCellLines = n_distinct(cellLine),
              nDrugs     = n_distinct(drug))
```

Let's also try something a little more complex.
We can also count the number of unique drug concentrations
**in each study** separately.

```{r count-conc}
## with base R
tapply(pharmacoData$concentration, pharmacoData$study,
       function(x) { length(unique(x)) })

## with the tidyverse
pharmacoData %>%
    group_by(study) %>%
    summarize(n = n_distinct(concentration))
```

One of the first things data scientists do when
digging into new data is to explore their distributions. 
Histograms visualize the data distributions and can also point us towards statistical 
models to use. The code below  transforms the concentration values to the logarithmic scale
and plots a histogram separately for each study. 

```{r conc-hist, fig.width = 8, fig.height = 4, warning = FALSE}
pharmacoData %>%
    ggplot(aes(x = log2(concentration))) +
    geom_histogram(fill = "gray", color = "black") +
    facet_wrap(~ study) +
    ggtitle("Distributions of concentrations by study")
```

Based on these plots, which study would you say has the most consistent
experimental protocol?

> Place your answer here

Viability scores are the percentage
of cells that survive upon exposure to a certain drug.
Below, we will explore the range of the data and calculate how
many data points are below 0 and above 100.

```{r viability-summaries}
## with base R
range(pharmacoData$viability)
sum(pharmacoData$viability < 0)
sum(pharmacoData$viability > 100)

## with the tidyverse
pharmacoData %>%
    summarize(min_viability = min(viability),
              max_viability = max(viability),
              n_too_small   = sum(viability < 0),
              n_too_big     = sum(viability > 100))
```

We can also compare the distribution of viability scores between 
the two studies using density plots.

```{r log-density, fig.width = 8, fig.height = 4, warning = FALSE}
pharmacoData %>%
    ggplot(aes(x = viability, group = study, fill = study, color = study)) +
    geom_density(alpha = 1/4) +
    xlim(0, 170) +
    ggtitle("Distributions of viability scores by study")
```

Based on the distribution of the viability scores, would you say there
are obvious differences between the two studies?

> Place your answer here

The code below plots the viability scores as box-plots for each drug,
stratified by the two studies. We highlight the region of the plot
where viability scores should fall (between 0 and 100).

```{r viability-boxplots, fig.width = 8, fig.height = 4, warning = FALSE}
gp <- pharmacoData %>%
    ggplot(aes(y = viability, x = drug, fill = study)) +
    scale_x_discrete() + 
    annotate(geom = "rect", ymin = 0, ymax = 100, xmin = -Inf, xmax = Inf,
             fill = 'black', alpha = 1/6) +
    geom_boxplot(outlier.alpha = 1/5) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1/2)) +
    ggtitle("Distributions of viability scores by drug and study")
gp
```

There appear to be a few outliers with incredibly high viability scores!
We should keep this in mind, but to get a better look at the majority of the
data, we can limit the y-axis of the plot.

```{r viability-boxplots-limit, fig.width = 8, fig.height = 4, warning = FALSE}
gp + ylim(0, 200)
```

Can you tell something about the toxic properties of the different 
drugs? Are these properties consistent across studies?

> Place your answer here
