---
title: "Supplement: Exploring High Dimensional Data with PCA and Clustering"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
```

## Principal Components Analysis

Principal Components Analysis (PCA) is often used for dimensionality reduction when dealing with high-dimensional data. Here, we will look at a subset of the drug sensitivity data, from 
the GDSC study. Each cell line will be treated as an independent observation with multiple
features corresponding to the AUC values for the interaction between the cell line and 
each drug in the study.

To start, let's load the data from the GDSC study and format it into a `data.frame` of AUC
values.

```{r}
sumData <- readRDS(file.path("..", "data", "summarizedPharmacoData.rds"))
auc_GDSC <- spread(sumData[,c(1,2,6)], drug, auc_GDSC)
rownames(auc_GDSC) <- auc_GDSC$cellLine
auc_GDSC$cellLine <- NULL
auc_GDSC <- auc_GDSC[!is.na(rowSums(auc_GDSC)), ]
head(auc_GDSC)
```

As we can see, we now have a `data.frame` of AUC values with rows corresponding to cell
lines and columns corresponding to drugs.

We are now ready to perform PCA, via the `prcomp` 
function. This function computes all of the relevant information and returns it as 
a `list`, which we will name `pca`. Note: this function expects the independent 
observations to be in the *rows* of the input.

```{r}
pca <- prcomp(auc_GDSC)
```

Now that the calculation is done, we can start to visualize the results. A reasonable
starting point is to plot the top 2 principal components (PCs) and see what
structure (if any) we can observe in the data.

```{r}
plot(pca$x, asp = 1)
```

For a more detailed picture, we may want to look at a larger number of PCs. We can
use the `pairs` function to plot larger numbers of dimensions against each other.

```{r}
pairs(pca$x[,1:5], asp = 1)
```

Another important consideration is the relative importance of the PCs. Calling the
`plot` function directly on the output of `prcomp` produces a barplot of the variances
of the PCs, which is one way of quantifying the amount of information they contain. 
Notice that the height of the bars is strictly decreasing; this will always be the case
for PCA.

```{r}
plot(pca)
```

Since the first PC seems to contain a fair amount of information, we may also be 
interested in examining which of the drugs are most related to it. We can do this
by plotting the loadings of each drug on PC-1, which are stored in the `rotation`
matrix.

```{r}
barplot(pca$rotation[,1], las = 2)
```


## Clustering

Another question we may be interested in answering is whether or not there are distinct
groups of cell lines, perhaps based on cancer type or tissue of origin. With no 
additional information on the cell lines, we can seek to discover such groups via
clustering.

### k-means

Below, we apply k-means clustering to the same subset of the data as above.

```{r}
km <- kmeans(auc_GDSC, centers = 3)
```

And we visualize the resulting clusters on the top two PCs:

```{r}
plot(pca$x, asp = 1, col = km$cluster)
```

Alternatively, we could have applied our clustering algorithm to a reduced 
representation of the data. Here, we will apply k-means to the top 3 PCs of the
data and again visualize the results using the top 2 PCs.

```{r}
km2 <- kmeans(pca$x[,1:3], centers = 3)
plot(pca$x, asp = 1, col = km2$cluster)
```

Why do these results differ from the clusters obtained on the full dataset?
(Hint: look at the `pairs` plot).


### Hierarchical clustering

We will try one more clustering method, this time using the `hclust` function to
perform hierarchical clustering.

```{r}
hc <- hclust(dist(auc_GDSC))
lab <- cutree(hc, k = 3)

plot(pca$x, asp = 1, col = lab)
```

As with k-means, we may also choose to cluster the low-dimensional representation
of our data rather than the full dataset:

```{r}
hc <- hclust(dist(pca$x[,1:3]))
lab <- cutree(hc, k = 3)

plot(pca$x, asp = 1, col = lab)
```


